Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-21 paper2.qmd
section:6, process:write

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods. In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. @sec-discussion discusses the findings, limitations, and challenges and explores avenues for future research. Finally, @sec-conclusion summarizes the study's conclusions.

@@

<!-- 3. But does it hold in practice? -->
Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data. In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Bramley_2008; @Kelly_et_al_2022], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics. For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]. These misfit texts may exhibit larger discriminal dispersions or represent outliers --texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969]. In either case, comparing misfit texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them.

@@

<!-- 5. So, what happens when the assumption does not hold? -->
Significant statistical and measurement issues can arise when the assumption of equal dispersions between stimuli does not hold. Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Furthermore, if researchers acknowledge that misfit statistics may represent outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.

@@

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory). Additionally, based on the arguments developed in @sec-theory-issue1b, the SCM incorporates the judges' biases $(B_{kj})$. Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$. Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:58:20 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->

<!-- make a summary of what you have accomplished so far -->

<!-- change the paradigm -->

<!-- opportunities open  -->

<!-- Now that we have this model we have a principled way to make some power analysis which will helps us to design studies, rather than just relying in simple rules of thumbs. -->

<!-- Thus, this study addresses the issues identified in @sec-theory-issues by extending  Thurstone’s general form using the structural approach to causal inference and Bayesian inference methods. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons, i.e, the sampleand compariosn mechanisms. In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis. -->

<!-- "If Thurstone’s Law is to be used in support of comparative judgment for assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes." [@Kelly_et_al_2022, pp. 678] -->

@@

@@

<!-- Pollitt (2012a) argues that assessors in pairwise comparisons need less training, given intuitive comparison and decision making. However, there is a distinct lack of research on the role of training. [?] -->

@@

@@

<!-- sampling and comparisons mechanism are effective generators of missing data. -->

@@

<!-- design matters -->

@@

@@

<!-- Due to efficiency of scoring, most of research only collects only one text. This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability. However, this cannot be known ad-hoc.  -->

@@

<!-- Brought Rubin's missing data framework to the field. -->
<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are are random samples, they fulfill ignorability.  -->

<!-- recording: Sven and Tine 25.04.10; time: 00:12:00 - 00:14:00 -->
<!-- Consider how to use the S and C, when the designs are more complexes. -->
<!-- Normally, S and C are not typically mentioned  -->

@@

## Limitations and further research {#sec-discussion-limitations}

>>>>

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines Causal and Bayesian inference methods. In addition to potentially enhancing measurement reliability and validity [@Perron_et_al_2015], and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key CJ experimental design features, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. @sec-discussion discusses the findings, limitations, and challenges and explores avenues for future research. Finally, @sec-conclusion summarizes the study's conclusions.

@@

<!-- 3. But does it hold in practice? -->
Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data. In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Bramley_2008; @Kelly_et_al_2022], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics. For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]. These misfit texts may exhibit larger discriminal dispersions or represent outliers--texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969]. In either case, the comparison of these texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them, except their elimination from the analysis.

@@

<!-- 5. So, what happens when the assumption does not hold? -->
Significant statistical and measurement issues can arise when the assumption of equal dispersions between stimuli does not hold. Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.

@@

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory). Additionally, building on the arguments developed in @sec-theory-issue1b, the SCM incorporates a key design features of CJ experiments: the judges' biases $(B_{kj})$. These variables, along with the outcome $(O_{iahbjk})$, constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$. Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

<!-- recording: Sven and Tine 24.11.25; time: 00:56:10 - 00:58:20 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:12:00 - 00:14:30 -->
<!-- recording: Sven and Tine 25.04.10; time: 00:41:55 - 00:47:00 -->

<!-- 1. General summary of what we have done. What are the first benefits? -->
This study extends Thurstone's general form through a systematic, integrated approach that combines Causal and Bayesian inference methods. It begins by formulating the extension as a conceptual model, represented by an SCM and a DAG. This formulation integrates Thurstone's core theoretical principles, including the distribution of the discriminal process, with key CJ experimental design features, such as the sampling and comparison mechanisms. The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model. Specifically, the statistical model accounts for judge biases, captures the hierarchical structure of stimuli, incorporates measurement error in trait estimation and hypothesis testing, and accommodates violations of the equal dispersion assumption. These improvements potentially enhance measurement reliability and validity [@Perron_et_al_2015] and improve the statistical accuracy of hypothesis testing.

<!-- 2. Are there any additional advantages? -->
Beyond these potential benefits, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data. Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs--one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.

<!-- If Thurstone's Law is to be used in support of comparative judgment for assessment, his theory must be extended to cover the uses and applications that are required for assessment purposes. [@Kelly_et_al_2022, pp. 678] -->

<!-- We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used. However, work to derive some general principles for good practice would be valuable. [@Kelly_et_al_2022, pp. 683] -->

<!-- 3. First research avenue -->
One promising research avenue involves investigating how the hierarchical structure of stimuli affects CJ experiment designs and hypotheses testing. One such study could examine whether collecting a single text per individual provides sufficient information to produce reliable trait estimates at different levels of the hierarchy. Due to concerns about the feasibility of the comparison task [@Boonen_et_al_2020], many CJ studies assess traits using only one text per individual [@citation]. This experimental design implicitly assumes that within-individual variability is small relative to between-individual differences--that is, texts produced by the same individual are relatively homogeneous. However, researchers rarely test this assumption, even though it directly influences decisions about the number of texts and participants needed for the experiment, ultimately affecting the reliability of the trait estimates.

Relatedly, another study could investigate hypotheses about the characteristics of the texts themselves. Previously, the literature lacked a direct and integrated method for evaluating these types of hypotheses. Now, by collecting data on specific stimulus features, researchers can assess whether these variables produce the expected effects.

<!-- power analysis to know the implications of certain sampling is nice to have it explicitly. -->

<!-- One of these avenues could be about the implications of the hierarchical structure of the stimuli, not only for the design of CJ experiments, but also for the test of hypothesis.  -->
<!-- The avenues of research span the following:  -->
<!-- - assessing if the number of texts a researcher collects is sufficient. Due to efficiency of scoring, most of research only collects only one text. This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability. However, this cannot be known ad-hoc. The frameworks open the avenue to test this assumption -->
<!-- - Assessing hypothesis regarding the characteristics of the texts. Before the literature did not have a direct and integrated way to assess these types of hypothesis, now you just need to collect data on the particular stimuli, and test if these variables reflect the expected effect -->

<!-- Derived from this other avenues of research that are open are -->
<!-- - assessing if the number of individuals a researcher collects is sufficient. Most of the literature focuses on ensuring that the number of comparisons between texts are sufficient, but no literature actively engages in determining if the number of individuals are sufficient to test hypothesis at that level. -->
<!-- - Moreover, if certain characteristics of the individuals explain the trait locations of their texts -->

@@

<!-- presence of judges biases in the comparative judgment procedure. From this it derives research on:  -->
<!-- - does the judges have biases, and if so, do ad-hoc sta -->

<!-- Given that the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, this study has argued that judges biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II].  -->

<!-- This hierarchical nature open paths for research in, for instance, the benefits of *repeated measures designs* [@Lawson_2015, pp. 366-376] in CJ experiments. This is when the experimental conditions accounts for judges comparing the same pair of stimuli multiple times. Or research that determines the minimum number of judges and judgments required to ensure an acceptable level of reliability and validity under particular assumptions of the data [@Kelly_et_al_2022, pp. 682]. -->

@@

<!-- The primary writing study showed that both trained markers and teachers with no marker training can use the method effectively to assess the quality of children’s writing, and the e-scape project has demonstrated the same for the assessment of e-portfolios in Design & Technology. We do not yet know how widely it can be, or should be, applied. [@Pollitt_2012b, pp. 297]-->

<!-- However, the definition of expert in this assessment context is further complicated by studies suggesting that training is unnecessary for comparative judgment exercises to be conducted well. [@Kelly_et_al_2022, pp. 681] -->

@@

<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait. --->

@@

@@

<!-- clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments. Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges. -->

<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments. Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges. It also accounts for potential judge biases in the comparative judgment process. -->

@@

<!-- Continuing with the first advantage, now we clarify two additional processes of CJ experiments that previous literature do not typically mention: that the sample and comparison mechanism are generators of missing data--that is, they determine which observations do we get to see in our data sample. This opens the path to use Rubin's missing data framework [@Rubin_1976] into CJ applications to investigate more complex CJ designs, particularly the ones that are not random samples as the ones outlined here, which could be characterized in Rubin's structure parlor as missing at random (MAR) and missing not at random (MNAR) [@Rubin_1976]. Among these cases are adaptive comparative algorithm designs, for which the causal structure is not clear, and thus, its impact on trait measurement and hypothesis testing is also not clear. Another  -->

<!-- normally S and C are not explicitly mentioned  -->

<!-- Brought Rubin's missing data framework to the field. -->
<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are random samples, they fulfill ignorability. -->

@@

@@

@@
